---
title: "DESeq / GSEA"
author: "Yi Jou (Ruby) Liao"
output: 
  rmarkdown::html_vignette:
    df_print: paged
    fig_width: 6
    fig_height: 4
    toc: true
    number_sections: true
bibliography: vignettes.bib
csl: elsevier-vancouver.csl
vignette: >
  %\VignetteIndexEntry{DESeq / GSEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Setup

```{r setup}
library(Rubrary)
library(dplyr)
library(tibble)

# For `airway` install
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)
```

# Introduction

The example dataset for this vignette comes from the paper [@himesRNASeqTranscriptomeProfiling2014] below:

> Himes et al. (2014). "RNA-Seq Transcriptome Profiling Identifies CRISPLD2 as a Glucocorticoid Responsive Gene that Modulates Cytokine Function in Airway Smooth Muscle Cells." PLoS ONE, 9(6), e99625. <https://doi.org/10.1371/journal.pone.0099625>.

From the abstract:

> Using RNA-Seq, a high-throughput sequencing method, we characterized transcriptomic changes in four primary human ASM cell lines that were treated with dexamethasone---a potent synthetic glucocorticoid (1 µM for 18 hours).

The experiment was performed on human airway smooth muscle (ASM) cells. Glucocorticoids are a commonly used drug to treat asthma for its anti-inflammatory effects in lung tissue. The goal of the experiment was to characterize the mechanism of glucocorticoid inflammation suppression through RNAseq. Himes et al. found a number of glucocorticoid-responsive genes such as DUSP1 and CRISPLD2. They also identified enrichment in functional annotation of terms such as "glycoprotein/extracellular matrix", "vasculature development", and "circulatory system process".

In the original paper, the authors used [Cuffdiff 2](https://www.nature.com/articles/nbt.2450) to find differential gene expression and used the [NIH Database for Annotation, Visualization, and Integrated Discovery (DAVID)](https://david.ncifcrf.gov/) for gene set enrichment analysis.

# Load data

The package `airway` contains data from the Himes et al. paper in the form of a RangedSummarizedExperiment object and is hosted on Bioconductor.

## Get `airway` data from package

```{r Load airway data, message=FALSE}
if (!requireNamespace("airway", quietly = TRUE)){
  BiocManager::install("airway")
}

library(airway)
data(airway)
```

## Get counts and sample annotation

The gene expression data is represented in counts. There are 4 samples treated with dexamethasone (`anno$dex == "trt"`) and 4 untreated samples (`anno$dex == "untrt"`).

```{r Get counts and annotation}
# Counts
cts <- assay(airway, "counts")
# Annotation
anno <- colData(airway) %>%
  as.data.frame()
anno$dex <- relevel(anno$dex, "untrt") # Set untreated as first factor

head(cts)
head(anno)
```

## Visualize samples w/ PCA

```{r Visualize samples w/ PCA}
# log2UQ
# Rubrary::run_PCA
# Rubrary::plot_PCA
```

# `DESeq2` differentially expressed genes

`DESeq2` [@loveModeratedEstimationFold2014] performs differential gene expression analysis my modeling counts as a negative binomial distribution.

## Create DESeqDataSet object and run DESeq

`Rubrary::output_DESeq` takes the DESeqResults object, transforms it to a dataframe, and adds a sign log2 p-value column.

```{r Create DESeqDataSet object and run DESeq}
library(DESeq2)

dds <- DESeqDataSetFromMatrix(
  countData = cts[,rownames(anno)],
  colData = anno,
  design = ~ dex)

dds <- DESeq(dds)
res <- results(dds, contrast = c("dex", "trt", "untrt"))
deseq_res_df <- Rubrary::output_DESeq(res)
```

### Replace Ensembl ID w/ HGNC symbol

`airway` gene names are given in [Ensembl ID](http://www.ensembl.org/info/genome/stable_ids/index.html), but [HGNC gene symbols](https://www.genenames.org/about/guidelines/) can be easier to interpret in downstream analysis and more compatible with gene set enrichment analysis (GSEA) tools.

```{r Replace Ensembl ID w/ HGNC symbol}
# Gene format is currently in Ensembl ID format
head(deseq_res_df$gene)

library(biomaRt)
# Load Mart object - database of Homo sapiens genes from Ensembl
mart <- biomaRt::useDataset(
  dataset = "hsapiens_gene_ensembl",
  mart = biomaRt::useMart("ENSEMBL_MART_ENSEMBL"))

conv <- Rubrary::convert_genes(
    genes = deseq_res_df$gene,
    from_to = c("ensembl_gene_id", "hgnc_symbol"),
    mart = mart, table = T) %>%
  filter(hgnc_symbol != "") %>% # Exclude unmatched HGNC symbols
  distinct(hgnc_symbol, .keep_all = T)

deseq_res_hgnc <- dplyr::left_join( # Join stats with HGNC symbol table
  conv, deseq_res_df, by = join_by(ensembl_gene_id == gene)) %>%
  arrange(desc(sign_log_p))

# Genes now have associated HGNC symbols
head(deseq_res_hgnc)
```

## Filter to protein coding genes

For ease of interpretation, the DE genes results can also be filtered to protein coding genes only. `Rubrary::filter_genes()` can filter with the list of protein coding genes retrieved with `Rubrary::get_PC_genes()`. Since we already created the `Mart` object when converting gene symbols, we can reuse it to avoid querying the BioMart database again and load protein coding genes efficiently.

```{r PC genes}
deseq_res_hgnc_PC <- Rubrary::filter_genes(
  df = deseq_res_hgnc,
  gene_col = "hgnc_symbol",
  genes_filt = Rubrary::get_PC_genes(mart)
)
```

## Volcano plot

DESeq results can be visualized through volcano plots. Volcano plots show that top and bottom differentially expressed genes when sorted by decreasing signed log p value present as outliers on the top right and top left.

```{r Volcano_plot, fig.height=12, fig.width=16}
# All DE genes
head(deseq_res_hgnc[order(deseq_res_hgnc$sign_log_p, decreasing = T),])
tail(deseq_res_hgnc[order(deseq_res_hgnc$sign_log_p, decreasing = T),])

Rubrary::plot_volcano(
  df_deg = deseq_res_hgnc,
  names = "hgnc_symbol",
  x = "log2FoldChange", y = "pvalue",
  FCcutoff = 1, pCutoff = 5e-04,
  title = "Airway Treated vs. Untreated DE Genes",
  xlab_high = "\U2191 in treated,\n\U2193 in untreated",
  xlab_low = "\U2191 in untreated,\n\U2193 in treated",
  subtitle = "All genes"
)

# Protein coding DE genes
head(deseq_res_hgnc_PC[order(deseq_res_hgnc$sign_log_p, decreasing = T),])
tail(deseq_res_hgnc_PC[order(deseq_res_hgnc$sign_log_p, decreasing = T),])

Rubrary::plot_volcano(
  df_deg = deseq_res_hgnc_PC,
  names = "hgnc_symbol",
  x = "log2FoldChange", y = "pvalue",
  FCcutoff = 1, pCutoff = 5e-04,
  title = "Airway Treated vs. Untreated DE Genes",
  xlab_high = "\U2191 in treated\n\U2193 in untreated",
  xlab_low = "\U2191 in untreated\n\U2193 in treated",
  subtitle = "Protein coding genes"
)
```

# `fgsea` gene set enrichment analysis

`fgsea` [@korotkevichFastGeneSet2016] is an implementation of an algorithm for fast gene set enrichment analysis (GSEA), based on the original GSEA algorithm [@subramanianGeneSetEnrichment2005].

## Get pathways from `msigdbr`

Gene set enrichment analysis relies on inputting gene sets / pathways related to biological function. The Molecular Signatures Database (MSigDB) [@liberzonMolecularSignaturesDatabase2011] was published as a database of annotated pathways to facilitate GSEA and is accessible through the R package `msigdbr` [@dolgalevMsigdbrMSigDBGene2022].

[Human MSigDB Collections](https://www.gsea-msigdb.org/gsea/msigdb/human/collection_details.jsp)

-   C2CP: curated gene sets, canonical pathways

-   C5GO: ontology gene sets, Gene Ontology gene sets

    -   BP: GO Biological Process ontology

    -   CC: GO Cellular Component ontology

    -   MF: GO Molecular Function ontology

-   H: hallmark gene sets [@liberzonMolecularSignaturesDatabase2015]

```{r Get pathways from `msigdbr`}
if (!requireNamespace("msigdbr", quietly = TRUE)){
  BiocManager::install("msigdbr")
}

msigdbr::msigdbr_collections()

pthwys <- rbind(
  msigdbr::msigdbr(category = "C2", subcategory = "CP"),
  msigdbr::msigdbr(category = "C5", subcategory = "GO:BP"),
  msigdbr::msigdbr(category = "C5", subcategory = "GO:CC"),
  msigdbr::msigdbr(category = "C5", subcategory = "GO:MF"),
  msigdbr::msigdbr(category = "H")
) %>%
  split(x = .$gene_symbol, f = .$gs_name) # Named list of pathways
```

## Run `fgsea`

```{r Run fgsea}
# `fgsea` requires a named numeric vector as input to the `stats` argument
deseq_stats <- setNames(
    deseq_res_hgnc_PC[,"sign_log_p"], 
    deseq_res_hgnc_PC[,"hgnc_symbol"]
)

gsea_results <- fgsea::fgsea(
  pathways = pthwys,
  stats = deseq_stats,
  eps = 0.0,
  minSize = 15,
  maxSize  = 500) %>%
  arrange(desc(NES)) %>%
  mutate(sign_log10_p = sign(NES) * -log10(pval)) # Create sign_log10_p column

```

## Visualize GSEA results as a table

See the [GSEA User Guide: Interpreting GSEA Results](https://www.gsea-msigdb.org/gsea/doc/GSEAUserGuideTEXT.htm#_Interpreting_GSEA_Results) for detailed explanation in interpreting results.

```{r GSEA_table, fig.width=12, fig.height=6}
# Get top significant pathways by positive/negative enrichment score (ES)
pos_pws <- gsea_results[ES > 0][head(order(pval), n = 5), pathway]
neg_pws <- gsea_results[ES < 0][head(order(pval), n = 5), pathway]
top_pws <- c(pos_pws, rev(neg_pws))

fgsea::plotGseaTable(
  pathways = pthwys[top_pws],
  stats = deseq_stats,
  fgseaRes = gsea_results,
  gseaParam = 0.5
)
```

## Pathway enrichment plot

Specific pathways of interest can be plotted in GSEA Java app-like enrichment plots that combine a waterfall and mountain plot.

The top enriched pathway from our GSEA results is "[HALLMARK_ADIPOGENESIS](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/HALLMARK_ADIPOGENESIS.html)", a collection of 196 genes up-regulated during adipocyte (fat cell) differentiation. This is in line with studies reporting the complex effects of glucocorticoids on adipose tissue biology, including differentiation of adipocyte precursors and adipogenesis [@haunerGlucocorticoidsInsulinPromote1987; @pantojaGlucocorticoidSignalingDefines2008].

The maximum absolute value in the Enrichment Score mountain plot corresponds to the `ES` value reported in the GSEA results.

```{r GSEA_pathway, fig.width=8, fig.height=6}
gsea_results[gsea_results$pathway == "HALLMARK_ADIPOGENESIS",]

Rubrary::plot_GSEA_pathway(
  sig = deseq_res_hgnc_PC,
  geneset = pthwys[["HALLMARK_ADIPOGENESIS"]],
  genecol = "hgnc_symbol",
  rankcol = "sign_log_p",
  rankcol_name = "DESeq sign log2 p-value",
  label = FALSE,
  title = "Airway DE Genes: HALLMARK_ADIPOGENESIS",
  lab_high = "\U2191 in treated\n\U2193 in untreated",
  lab_low = "\U2191 in untreated\n\U2193 in treated"
)
```

For gene sets of smaller sizes, like 20 gene pathway "[GOBP_NEURON_FATE_SPECIFICATION](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/GOBP_NEURON_FATE_SPECIFICATION.html)", labeling each gene may be appropriate.

```{r GSEA_pathway_labeled, fig.width=8, fig.height=6}
gsea_results[gsea_results$pathway == "GOBP_NEURON_FATE_SPECIFICATION",]

Rubrary::plot_GSEA_pathway(
  sig = deseq_res_hgnc_PC,
  geneset = pthwys[["GOBP_NEURON_FATE_SPECIFICATION"]],
  genecol = "hgnc_symbol",
  rankcol = "sign_log_p",
  rankcol_name = "DESeq sign log2 p-value",
  label = TRUE,
  title = "Airway DE Genes: GOBP_NEURON_FATE_SPECIFICATION",
  lab_high = "\U2191 in treated\n\U2193 in untreated",
  lab_low = "\U2191 in untreated\n\U2193 in treated"
)
```

Multiple gene set enrichment plots can be created in a batch with `Rubrary::plot_GSEA_batch`, a wrapper for `Rubrary::plot_GSEA_pathway` that works well with `lapply`. We can use the `neg_pws` list of pathway names to plot gene enrichment plots for the most significant negatively enriched pathways.

```{r GSEA_pathway_batch, fig.width=8, fig.height=6}
gsea_results[gsea_results$pathway %in% neg_pws,]

lapply(
  neg_pws[1:3],
  FUN = Rubrary::plot_GSEA_batch,
  genecol = "hgnc_symbol",
  pthwys = pthwys,
  sig = deseq_res_hgnc_PC,
  rankcol = "sign_log_p",
  rankcol_name = "DESeq sign log2 p-value",
  hllab = "Pathway genes",
  lab_high = "\U2191 in treated\n\U2193 in untreated",
  lab_low = "\U2191 in untreated\n\U2193 in treated"
)

```

# Citations
