---
title: "Gene Set Enrichment Analysis (GSEA)"
output: 
  html_document:
    df_print: "paged"
    fig_width: 6
    fig_height: 4
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
bibliography: vignettes.bib
csl: elsevier-vancouver.csl
vignette: >
  %\VignetteIndexEntry{Gene Set Enrichment Analysis (GSEA)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

```{r setup}
library(Rubrary)
library(dplyr)
library(tibble)
```

This is a continuation of the ["Differential Gene Expression" vignette](https://www.liaoyjruby.com/Rubrary/articles/DESeq_GSEA.html) and loads in the `airway` DESeq2 results to further investigate gene sets.

```{r airway results}
data("airway_deseq_res")
rmarkdown::paged_table(airway_deseq_res)
```

# `fgsea` Gene Set Enrichment Analysis

`fgsea` [@korotkevichFastGeneSet2016] is an implementation of an algorithm for fast gene set enrichment analysis (GSEA), based on the original GSEA algorithm [@subramanianGeneSetEnrichment2005].

## Get pathways from `msigdbr`

Gene set enrichment analysis relies on inputting gene sets / pathways related to biological function. The Molecular Signatures Database (MSigDB) [@liberzonMolecularSignaturesDatabase2011] was published as a database of annotated pathways to facilitate GSEA and is accessible through the R package `msigdbr` [@dolgalevMsigdbrMSigDBGene2022].

[Human MSigDB Collections](https://www.gsea-msigdb.org/gsea/msigdb/human/collection_details.jsp)

-   C2CP: curated gene sets, canonical pathways

-   C5GO: ontology gene sets, Gene Ontology gene sets

    -   BP: GO Biological Process ontology

    -   CC: GO Cellular Component ontology

    -   MF: GO Molecular Function ontology

-   H: hallmark gene sets [@liberzonMolecularSignaturesDatabase2015]

```{r Get pathways from `msigdbr`}
if (!requireNamespace("msigdbr", quietly = TRUE)){
  BiocManager::install("msigdbr")
}

rmarkdown::paged_table(msigdbr::msigdbr_collections())

pthwys <- rbind(
  msigdbr::msigdbr(category = "C2", subcategory = "CP"),
  msigdbr::msigdbr(category = "C5", subcategory = "GO:BP"),
  msigdbr::msigdbr(category = "C5", subcategory = "GO:CC"),
  msigdbr::msigdbr(category = "C5", subcategory = "GO:MF"),
  msigdbr::msigdbr(category = "H")
) %>%
  split(x = .$gene_symbol, f = .$gs_name) # Named list of pathways
```

## Run `fgsea`

See the [GSEA User Guide: Interpreting GSEA Results](https://www.gsea-msigdb.org/gsea/doc/GSEAUserGuideTEXT.htm#_Interpreting_GSEA_Results) for detailed explanation in interpreting results.

```{r Run fgsea}
# `fgsea` requires a named numeric vector as input to the `stats` argument
deseq_stats <- setNames(
    airway_deseq_res[,"sign_log_p"], 
    airway_deseq_res[,"hgnc_symbol"]
)

gsea_results <- fgsea::fgsea(
  pathways = pthwys,
  stats = deseq_stats,
  eps = 0.0,
  minSize = 15,
  maxSize  = 500) %>%
  arrange(desc(NES)) %>%
  mutate(sign_log10_p = sign(NES) * -log10(pval)) # Create sign_log10_p column

rmarkdown::paged_table(gsea_results)
```

## Visualize GSEA results as a table

Using `fgsea`'s `plotGseaTable` function.

```{r GSEA_table, fig.width=12, fig.height=6}
# Get top significant pathways by positive/negative enrichment score (ES)
pos_pws <- gsea_results[ES > 0][head(order(pval), n = 5), pathway]
neg_pws <- gsea_results[ES < 0][head(order(pval), n = 5), pathway]
top_pws <- c(pos_pws, rev(neg_pws))

fgsea::plotGseaTable(
  pathways = pthwys[top_pws],
  stats = deseq_stats,
  fgseaRes = gsea_results,
  gseaParam = 0.5
)
```

Using `Rubrary::plot_GSEA_barplot` function - simpler visualization of GSEA NES values but a bit nicer looking?

```{r plot_GSEA_barplot, fig.height=5, fig.width=10}
Rubrary::plot_GSEA_barplot(
  gsea_res = gsea_results,
  n_pws = 5, NES_cutoff = 1.5,
  pw_format = TRUE,
  cols = c("red", "blue"),
  title = "Airway GSEA"
)
```


## Pathway enrichment plot

Specific pathways of interest can be plotted in GSEA Java app-like enrichment plots that combine a waterfall and mountain plot.

The top enriched pathway from our GSEA results is "[HALLMARK_ADIPOGENESIS](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/HALLMARK_ADIPOGENESIS.html)", a collection of 196 genes up-regulated during adipocyte (fat cell) differentiation. This is in line with studies reporting the complex effects of glucocorticoids on adipose tissue biology, including differentiation of adipocyte precursors and adipogenesis [@haunerGlucocorticoidsInsulinPromote1987; @pantojaGlucocorticoidSignalingDefines2008].

The maximum absolute value in the Enrichment Score mountain plot corresponds to the `ES` value reported in the GSEA results.

```{r GSEA_pathway, fig.width=9, fig.height=7}
rmarkdown::paged_table(
  gsea_results[gsea_results$pathway == "HALLMARK_ADIPOGENESIS",])

Rubrary::plot_GSEA_pathway(
  sig = airway_deseq_res,
  geneset = pthwys[["HALLMARK_ADIPOGENESIS"]],
  genecol = "hgnc_symbol",
  rankcol = "sign_log_p",
  rankcol_name = "DESeq sign log2 p-value",
  label = FALSE,
  title = "Airway DE Genes: HALLMARK_ADIPOGENESIS",
  lab_high = "\U2191 in treated\n\U2193 in untreated",
  lab_low = "\U2191 in untreated\n\U2193 in treated"
)
```

For smaller gene sets, like 20-gene pathway "[GOBP_NEURON_FATE_SPECIFICATION](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/GOBP_NEURON_FATE_SPECIFICATION.html)", labeling each gene may be appropriate.

```{r GSEA_pathway_labeled, fig.width=8, fig.height=6}
rmarkdown::paged_table(
  gsea_results[gsea_results$pathway == "GOBP_NEURON_FATE_SPECIFICATION",])

Rubrary::plot_GSEA_pathway(
  sig = airway_deseq_res,
  geneset = pthwys[["GOBP_NEURON_FATE_SPECIFICATION"]],
  genecol = "hgnc_symbol",
  rankcol = "sign_log_p",
  rankcol_name = "DESeq sign log2 p-value",
  label = TRUE,
  title = "Airway DE Genes: GOBP_NEURON_FATE_SPECIFICATION",
  lab_high = "\U2191 in treated\n\U2193 in untreated",
  lab_low = "\U2191 in untreated\n\U2193 in treated"
)
```

Multiple gene set enrichment plots can be created in a batch with `Rubrary::plot_GSEA_batch`, a wrapper for `Rubrary::plot_GSEA_pathway` that works well with `lapply`. We can use the `neg_pws` list of pathway names to plot gene enrichment plots for the most significant negatively enriched pathways.

```{r GSEA_pathway_batch_tbl, fig.width=8, fig.height=6}
rmarkdown::paged_table(
  gsea_results[gsea_results$pathway %in% neg_pws[1:3],])
```

```{r GSEA_pathway_batch, fig.width=8, fig.height=6, message=FALSE}
lapply(
  neg_pws[1:3],
  FUN = Rubrary::plot_GSEA_pathway_batch,
  genecol = "hgnc_symbol",
  pthwys = pthwys,
  sig = airway_deseq_res,
  rankcol = "sign_log_p",
  rankcol_name = "DESeq sign log2 p-value",
  hllab = "Pathway genes",
  lab_high = "\U2191 in treated\n\U2193 in untreated",
  lab_low = "\U2191 in untreated\n\U2193 in treated"
)
```

# Resources

# Citations
