---
title: "Principal Component Analysis (PCA) Quickstart"
author: "Yi Jou (Ruby) Liao"
output: 
  html_document:
    df_print: "paged"
    fig_width: 6
    fig_height: 4
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Principal Components Analysis (PCA) Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png"
)
```

```{r setup, message = FALSE}
library(Rubrary)
library(dplyr)

# For `palmerpenguins` install
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)
```

# Data

Using `palmerpenguins` dataset [from Allison Horst](https://allisonhorst.github.io/palmerpenguins/).

## Load data

Get `palmerpenguins` data and add a unique ID per entry.
```{r Get_ data, message=FALSE}
if (!requireNamespace("palmerpenguins", quietly = TRUE)){
  install.packages("palmerpenguins")
}
library(palmerpenguins)

penguins <- penguins %>%
  as.data.frame() %>%
  mutate(Sample_ID = paste0(penguins_raw$`Individual ID`, "_", penguins_raw$`Sample Number`))
  
rownames(penguins) <- penguins$Sample_ID

rmarkdown::paged_table(penguins)
```

# Run PCA

Subset data to numeric only, then run PCA.
```{r Run_PCA}
num <- penguins %>%
  dplyr::select(!c(species, island, sex, year, Sample_ID)) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::select(where(~!all(is.na(.x)))) # Filter out NAs
  
pca <- Rubrary::run_PCA(
  df = num,
  summary = T, tol = 0,
  center = T, scale = T,
  screeplot = T
)
```

# Plot PCA

Use resulting `prcomp` object to plot PCA scores while annotating by `species` metadata.
```{r Plot_PCA_species_anno}
Rubrary::plot_PCA(
  df_pca = pca,
  anno = penguins,
  annoname = "Sample_ID",
  annotype = "species",
  title = "Palmer Penguins PCA Scores - Species",
  label = FALSE,
  density = TRUE
)
```

Same PCA scores plot, but annotating by `sex` metadata.
```{r Plot_PCA_sex_anno}
Rubrary::plot_PCA(
  df_pca = pca,
  anno = penguins,
  annoname = "Sample_ID",
  annotype = "sex",
  title = "Palmer Penguins PCA Scores - Sex",
  label = FALSE,
  density = TRUE
)
```

Plotting as biplot to view both (standardized) PCs and loadings on same plot.
```{r Plot_PCA_biplot}
Rubrary::plot_PCA_biplot(
  obj = pca,
  anno = penguins,
  annoname = "Sample_ID",
  annotype = "species",
  title = "Palmer Penguins PCA Biplot - Species",
  label = "Loadings"
)
```

# Varimax rotation
```{r}
pca_vm <- Rubrary::rotate_varimax(pca)

# Visualize loadings
Rubrary::plot_PCA(
  df_pca = pca_vm$loadings,
  PCx = "V1", PCy = "V2",
  title = "Palmer Penguins PCA Loadings (Varimax)",
  type = "Loadings",
  label = TRUE
)
```


```{r}
# Visualize scores
Rubrary::plot_PCA(
  df_pca = pca_vm$scores,
  PCx = "V1", PCy = "V2",
  anno = penguins,
  annoname = "Sample_ID",
  annotype = "species",
  title = "Palmer Penguins PCA Scores (Varimax) - Species",
  label = FALSE,
  density = TRUE,
  type = "Scores"
)
```


