---
title: "Principal Component Analysis (PCA) Quickstart"
author: "Yi Jou (Ruby) Liao"
output: 
  html_document:
    df_print: "paged"
    fig_width: 6
    fig_height: 4
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
bibliography: vignettes.bib
vignette: >
  %\VignetteIndexEntry{Principal Components Analysis (PCA) Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png"
)
```

```{r setup, message = FALSE}
library(Rubrary)
library(dplyr)

# For `palmerpenguins` install
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)
```

This vignette goes over how to run the PCA related functions in Rubrary for dimensional reduction and visualization. For a more math-y, in-depth step by step walkthrough with lengthier explanations see the [PCA Walkthrough vignette](https://www.liaoyjruby.com/Rubrary/articles/PCA_Walkthrough.html).

# Beltran et al. 2016 PCA

This replicates the PCA portions of the [graeberlab-ucla/glab.library PCA tutorial](https://github.com/graeberlab-ucla/glab.library/blob/master/vignettes/PCA_tutorial/PCA_tutorial.Rmd) using [Beltran et al. 2016](https://www.nature.com/articles/nm.4045)[@beltranDivergentClonalEvolution2016] CRPC/NEPC data that initiate many new computational Graeber Lab members into the lab.

## Load data

Raw text files can be loaded through `read.delim` by passing in the URL.

```{r Load Beltran data}
# Upper quartile normalized log2 transformed gene counts
Beltran_log2UQ_exp <- Rubrary::rread("https://raw.githubusercontent.com/graeberlab-ucla/glab.library/master/vignettes/PCA_tutorial/Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt", row.names = 1)

rmarkdown::paged_table(Rubrary::corner(Beltran_log2UQ_exp))
```

## Run PCA

Generally, scaling is unnecessary for log transformed gene expression because all features (genes) have the same unit of measure and it is assumed that scales are relatively similar across all genes.

Centering should always be performed for PCA to result in maximizing variance.

See the [Standardization section in the PCA Walkthrough](https://www.liaoyjruby.com/Rubrary/articles/PCA_Walkthrough.html#standardization) for more info.
```{r Beltran_Run_PCA}
Beltran_PCA <- Rubrary::run_PCA(
  df = Beltran_log2UQ_exp,
  center = TRUE, scale = FALSE
)
```

## Plot PCA

The annotations are in a huge file associating a large number of samples with its cancer type abbreviation. In this dataset, the samples correspond to "CRPC", castration resistant prostate cancer, and "NEPC", neuroendocrine prostate cancer.

```{r Load Beltran annotations}
# Annotation file
Beltran_info <- read.delim("https://raw.githubusercontent.com/graeberlab-ucla/glab.library/master/vignettes/PCA_tutorial/human.info.rsem.expression.txt")

head(Beltran_info)
```

The NEPC and CRPC samples separate pretty nicely along PC2.
```{r Beltran_Plot_PCA, fig.width=7, fig.height=5}
Rubrary::plot_PCA(
  df_pca = Beltran_PCA,
  type = "Scores",
  anno = Beltran_info,
  annoname = "sample",
  annotype = "type",
  density = T, ellipse = T
)
```

Lower PCs can be visualized simultaneously with `Rubrary::plot_PCA_matrix`.
```{r Beltran_PCA_mtx}
Rubrary::plot_PCA_matrix(
  df_pca = Beltran_PCA,
  PCs = c(1:5),
  anno = Beltran_info,
  annoname = "sample",
  annotype = "type"
)
```
3D view of the first three PCs.
```{r Beltran_PCA_3D}
Rubrary::plot_PCA_3D(
  df_pca = Beltran_PCA,
  PCs = c(1:3),
  anno = Beltran_info,
  annoname = "sample",
  annotype = "type"
)
```

## Varimax rotation

```{r Beltran varimax loadings}
Beltran_PCA_VM <- Rubrary::rotate_varimax(Beltran_PCA, normalize = FALSE)

# Visualize scores
Rubrary::plot_PCA(
  df_pca = Beltran_PCA_VM$scores,
  PCx = "V1", PCy = "V2",
  title = "Beltran 2016 PCA Scores (Varimax)",
  type = "Scores",
  anno = Beltran_info,
  annoname = "sample",
  annotype = "type",
  density = T, ellipse = T
)
```

# Palmer Penguins PCA

Using `palmerpenguins` dataset [from Allison Horst](https://allisonhorst.github.io/palmerpenguins/).

## Load data

Get `palmerpenguins` data and add a unique ID per entry.

```{r Get_ data, message=FALSE}
if (!requireNamespace("palmerpenguins", quietly = TRUE)){
  install.packages("palmerpenguins")
}
library(palmerpenguins)

penguins <- penguins %>%
  as.data.frame() %>%
  mutate(Sample_ID = paste0(penguins_raw$`Individual ID`, "_", penguins_raw$`Sample Number`))
  
rownames(penguins) <- penguins$Sample_ID

rmarkdown::paged_table(penguins)
```

## Run PCA

Subset data to numeric only, then run PCA.

```{r Run_PCA}
num <- penguins %>%
  dplyr::select(!c(species, island, sex, year, Sample_ID)) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::select(where(~!all(is.na(.x)))) # Filter out NAs
  
pca <- Rubrary::run_PCA(
  df = num,
  summary = T, tol = 0,
  center = T, scale = T,
  screeplot = T
)
```

## Plot PCA

Use resulting `prcomp` object to plot PCA scores while annotating by `species` metadata.

```{r Plot_PCA_species_anno}
Rubrary::plot_PCA(
  df_pca = pca,
  anno = penguins,
  annoname = "Sample_ID",
  annotype = "species",
  title = "Palmer Penguins PCA Scores - Species",
  label = FALSE,
  density = TRUE
)
```

Same PCA scores plot, but annotating by `sex` metadata.

```{r Plot_PCA_sex_anno}
Rubrary::plot_PCA(
  df_pca = pca,
  anno = penguins,
  annoname = "Sample_ID",
  annotype = "sex",
  title = "Palmer Penguins PCA Scores - Sex",
  label = FALSE,
  density = TRUE
)
```

Plotting as biplot to view both (standardized) PCs and loadings on same plot.

```{r Plot_PCA_biplot}
Rubrary::plot_PCA_biplot(
  obj = pca,
  anno = penguins,
  annoname = "Sample_ID",
  annotype = "species",
  title = "Palmer Penguins PCA Biplot - Species",
  label = "Loadings"
)
```

Plot multiple PCs simultaneously.
```{r Penguin_PCA_mtx}
Rubrary::plot_PCA_matrix(
  df_pca = pca,
  PCs = c(1:3),
  anno = penguins,
  annoname = "Sample_ID",
  annotype = "species",
)
```
3D view of the first three PCs.
```{r Penguin_PCA_3D}
Rubrary::plot_PCA_3D(
  df_pca = pca,
  PCs = c(1:3),
  anno = penguins,
  annoname = "Sample_ID",
  annotype = "species",
)
```


## Varimax rotation

```{r Varimax loadings}
pca_vm <- Rubrary::rotate_varimax(pca)

# Visualize loadings
Rubrary::plot_PCA(
  df_pca = pca_vm$loadings,
  PCx = "V1", PCy = "V2",
  title = "Palmer Penguins PCA Loadings (Varimax)",
  type = "Loadings",
  label = TRUE
)
```

```{r Varimax scores}
# Visualize scores
Rubrary::plot_PCA(
  df_pca = pca_vm$scores,
  PCx = "V1", PCy = "V2",
  anno = penguins,
  annoname = "Sample_ID",
  annotype = "species",
  title = "Palmer Penguins PCA Scores (Varimax) - Species",
  label = FALSE,
  density = TRUE,
  type = "Scores"
)
```

# References
